SNP Analysis

### 2019-02-05 Insilico Splits.Rmd
add Seattle SNP analysis scripts
Updating paper related scripts

```{r setup, include=FALSE}
library(ggplot2)
library(ggrepel)
library(pander)
library(reshape2)
library(gridExtra)
library(plyr) 
library(dplyr)
library(pROC)

@@

data = '~/Data/Ellie'

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
analysis.files = list.files(paste(data, 'Analysis', sep='/'), full.names=T)

@@

cache.dir = paste(data, 'Analysis/5e6_arms_all_logR', sep='/')
split.dir = paste(data, 'Analysis/5e6_arms_splits_logR', sep='/')

>>>>

```{r setup, include=FALSE}
library(tidyverse)
library(ggrepel)
library(pander)
library(reshape2)
library(gridExtra)
library(pROC)

@@

data = '~/Data/BarrettsProgressionRisk'

data.files = list.files(paste(data, 'QDNAseq/training',sep='/'), full.names=T)
analysis.files = list.files(paste(data, 'Analysis', sep='/'), full.names=T)

@@

cache.dir = paste(data, 'Analysis/5e6_arms_all', sep='/')
split.dir = paste(data, 'Analysis/5e6_arms_splits', sep='/')

### 2019-02-05 Paper.Rmd
Updating paper related scripts

```{r setup, include=FALSE}
library(biomaRt)
library(GenomicRanges)
library(stringr)
library(ggplot2)
library(ggrepel)
library(GGally)
library(plyr) 
library(pander)
library(reshape2)
library(gridExtra)
library(CoxHD)
library(plyr)
library(dplyr)

source('lib/load_patient_metadata.R')
source('lib/cv-pt-glm.R')
source('lib/common_plots.R')
source('lib/data_func.R')

data = '~/Data/Ellie'
info.dir = '~/Data/Ellie/QDNAseq'
#tile.w=c(5e06, 'arms')
#tile.w=5e+06
#tile.w=1e7
#tile.w='arms'

#includeDemo = F

@@

all.patient.info = read.patient.info(patient.file, demo.file, set='All')$info
head(all.patient.info)

patient.info = all.patient.info

validation.patient.info =  subset(all.patient.info, Set == 'Test')
validation.patient.info = arrange(validation.patient.info, Status, Hospital.Research.ID, Endoscopy.Year, Pathology)
sum.validation.info = summarise.patient.info(validation.patient.info)

@@

all.info = rbind(sum.patient.data, sum.validation.info)

@@

pander(
  as.data.frame(all.info %>% summarise(
  'Total Patients'=length(unique(Patient)),
  'Total Samples'=sum(total.samples),
  'Range'=paste(range(end.year-start.year), collapse='-'),
  'Median Years'=paste(median(end.year-start.year),'+/-',round(sd(end.year-start.year),1)),
  'Median Endoscopies'=paste(median(total.endos),'+/-',round(sd(total.endos),1)),
  'Median age'=paste(median(age.diagnosis),'+/-',round(sd(age.diagnosis),1)),
  'gender F:M'=paste(table(gender),collapse=':')
))
, caption='full cohort')
```

```{r echo=F}
cohortSummary = as.data.frame(all.info %>% dplyr::group_by(Status) %>% summarise(
  'Total Patients'=length(unique(Patient)),
  'Total Samples'=sum(total.samples),
  'Range follow-up (years)'=paste(range(end.year-start.year), collapse='-'),
  'Mean follow-up (years)'=paste(round(mean(end.year-start.year), 1),'+/-',round(sd(end.year-start.year),1)),
  'Range endoscopies'=paste(range(total.endos), collapse='-'),
  'Mean Endoscopies'=paste(round(mean(total.endos), 1),'+/-',round(sd(total.endos),1)),
  'Mean age'=paste(round(mean(age.diagnosis), 1),'+/-',round(sd(age.diagnosis),1)),
  'gender F:M'=paste(table(gender),collapse=':'),
  'Mean Length'=paste(round(mean(C,na.rm=T), 1),'+/-',round(sd(C,na.rm=T),1)),
  'naC'=length(which(is.na(C)))
))

cohortSummary = cbind(cohortSummary,
  rbind(table(subset(all.patient.info, Status == 'NP')$Pathology),
        table(subset(all.patient.info, Status == 'P')$Pathology)))

  rbind(table(subset(patient.info, Status == 'NP')$Pathology),
        table(subset(patient.info, Status == 'P')$Pathology))

  rbind(table(subset(validation.patient.info, Status == 'NP')$Pathology),
        table(subset(validation.patient.info, Status == 'P')$Pathology))

@@

```{r labelsPNP, echo=F, message=F, include=T}
cache.dir = paste(data, 'Analysis/5e6_arms_all', sep='/')

@@

@@

noHGD.pred.confidence = as.data.frame.matrix(table(pg.sampNOHGD$quants, pg.sampNOHGD$Status))
noHGD.pred.confidence = cbind(noHGD.pred.confidence, pg.sampNOHGD %>% dplyr::group_by(quants) %>% summarise ( 'mn'=mean(Prediction), 'sd'=sd(Prediction) ))
noHGD.pred.confidence$quant = rownames(noHGD.pred.confidence)

@@

ggplot(noHGD.pred.confidence, aes(mn, perc)) + 
  geom_rect(aes(xmin=r1, xmax=r2, ymin=0,ymax=1, fill=Risk), alpha=0.6) + scale_fill_manual(values=riskCols) +
  geom_point() + geom_text(aes(label=paste(r1,r2,sep='-')), nudge_x=0.05) +
  geom_errorbar(aes(ymin=ci.low, ymax=ci.high), width=0.01) + labs(x='Mean prediction', y='Ratio of `P` patients', title='Prediction calibration', subtitle='HGD excluded model') + plot.theme + theme(legend.position = 'none')

@@

  hgd.samples = subset(patient.info, Pathology %in% c('HGD','IMC'))
  hgd.df = dysplasia.df[hgd.samples$Samplename,]

  pred = predict(fitNOHGD, newx=hgd.df, s=lambda.opt, type='response')
  hgd.samples$Prediction = pred

  hgd.pred = hgd.samples %>% dplyr::group_by(Patient, Hospital.Research.ID, Status, Endoscopy.Year) %>% 
    summarise( max=max(Pathology), pred=Prediction[which.max(Pathology)] )

  rm(no.hgd.plots, coefs, performance.at.1se, models)
}

#hist(hgd.samples$Prediction)

@@

@@

@@

pred.confidence = qt %>% as.data.frame.matrix %>% dplyr::group_by(quant) %>% dplyr::mutate( 
  'perc'=P/sum(NP,P), 'p.value'=ft.fun(NP,P)$p.value, 'odds'=ft.fun(NP,P)$odds.ratio, 'conf'=ifelse(p.value < 0.05, '*', '') )

@@

pred.confidence$Risk = 'Moderate'
pred.confidence$Risk[ which(with(pred.confidence, p.value < 0.05 & perc < .5)) ] = 'Low'
pred.confidence$Risk[ which(with(pred.confidence, p.value < 0.05 & perc > .5)) ] = 'High'

pred.confidence = bind_cols(pred.confidence, 
                            data.frame(ci.low=qbeta(0.025, shape1=pred.confidence$P+.5, shape2 = pred.confidence$NP+.5),
                                      ci.high=qbeta(0.975, shape1=pred.confidence$P+.5, shape2 = pred.confidence$NP+.5)))

@@

```{r pcalib, echo=F, warning=F, message=F, fig.width=8, fig.height=6}
pcal = ggplot(pred.confidence, aes(mn, perc)) + 
  geom_rect(aes(xmin=r1, xmax=r2, ymin=0,ymax=1, fill=Risk), alpha=0.6) + scale_fill_manual(values=riskCols) +
  geom_point() + #geom_text(aes(label=paste(r1,r2,sep='-')), nudge_x=0.05) +
  geom_errorbar(aes(ymin=ci.low, ymax=ci.high), width=0.01) + labs(x='Mean prediction', y='Ratio of `P` patients', title='Prediction calibration') +
  scale_color_manual(values=riskCols) + plot.theme + theme(legend.position = 'none', text=element_text(size=10))
pcal

ggsave(filename='plots/pred_calib.png', plot=pcal, width=4, height=4, units='in', dpi=300)
```

@@

```{r indP, echo=F, warning=F, message=F, fig.width=5, fig.height=5}
back = merge(back, pred.confidence[c('quant','Risk')], by.x='quants',by.y='quant')
back = back %>% dplyr::group_by(Risk) %>% dplyr::mutate('se(Risk)'=(sd(Prediction)/sqrt(length(Risk)))*1.96 )

ggplot(back, aes(1:nrow(back), Prediction, color=Risk)) + 
  geom_point(aes(color=Risk, shape=Status)) +scale_color_manual(values=riskCols, name='P(P)') + 
  geom_hline(data=annotation.table, aes(yintercept=max), color='grey') + plot.theme +
  labs(x='sample', y='model probability', title='Per-sample probabilities')
```

```{r relrisk, echo=F, fig.width=6, fig.height=6}
#myPal = c(RColorBrewer::brewer.pal(9, 'Blues')[9:7], 
#          adjustcolor(grDevices::heat.colors(10)[6], blue.f=1.5),
#          ( grDevices::heat.colors(10)[7:6]),
#          RColorBrewer::brewer.pal(9, 'Reds')[6:9])

myPal = rev(RColorBrewer::brewer.pal(11, 'RdYlBu'))

unadjRR = ggplot(back, aes(OR)) + geom_histogram(aes(fill=..x..), bins=10, show.legend = F) +
  #scale_fill_distiller(palette = 'RdYlBu', type='div',name='P(P)') +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(y='n Samples', x='Relative Risk', title='Unadjusted relative risk') + plot.theme
unadjRR

ggplot(back, aes(Status, OR, color=OR)) + 
  geom_boxplot(outlier.colour=NA) + geom_jitter(width=0.3, aes(shape=Pathology)) + 
  scale_color_gradientn(colors = myPal,  name='') + plot.theme

@@

predhist = ggplot(back, aes(Prediction)) + geom_histogram(aes(fill=..x..), breaks=cuts, show.legend = F) +
    scale_fill_gradientn(colors = myPal,  name='') + 
  #scale_fill_distiller(palette = 'RdYlBu', type='div',name='P(P)') +
    plot.theme + labs(title='Predictions, all samples model', y='n Samples', x='Probability') 
predhist

ggsave(filename='plots/pred_hist.png', plot=predhist, width=5, height=5, units='in', dpi=300)

#ggplot(back, aes(OR, Prediction, color=Prediction)) + geom_point(show.legend = F) + scale_color_gradientn(colors = myPal,  name='')  + plot.theme
```

@@

el = subset(df2, early | late)
el2 = el %>% dplyr::group_by(Status, Patient) %>% dplyr::summarise( diff(arms), diff(CNA) )
wtA = round(wilcox.test(subset(el2, Status == 'Progressor')$`diff(arms)`, subset(el2, Status == 'Non-Progressor')$`diff(arms)`)$p.value, 3)
if (wtA < 0.0001) wtA = '< 0.0001'
#wilcox.test(subset(el2, Status == 'Progressor')$`diff(CNA)`, subset(el2, Status == 'Non-Progressor')$`diff(CNA)`)

@@

```{r jacknife, echo=T, warning=F, message=F, fig.height=10, fig.width=12}

coefs.per.pt = data.frame(matrix(ncol=length(pg.samp), nrow=nrow(ch), dimnames=list( ch$label, names(pg.samp))))
for (pt in names(pg.samp)) {
  lcf = loo.coefs[[pt]][,1]
  names(lcf) = rownames( loo.coefs[[pt]] )
  sharedcfs = intersect(names(lcf), ch$label)
  coefs.per.pt[sharedcfs, pt] = lcf[sharedcfs]
}
coefs.per.pt[is.na(coefs.per.pt)] = 0

@@

@@

# in this case jack.se is the SE of the variance 
# per coefficient
ch$jack.se = jack =  apply(coefs.per.pt, 1, function(x) {
  jk = jackknife(x, var)
  jk$jack.se
})
ggplot(ch, aes(reorder(label, coef), coef)) + geom_point() + 
  geom_errorbar(aes(ymin=coef-jack.se, ymax=coef+jack.se)) + 
  coord_flip() +
  plot.theme + labs(x='', title='Jackknifed coefficient CI')

@@

ggsave(filename='plots/pred_by_path.png', plot=rp, width=9, height=7, units='in', dpi=300)

>>>>

```{r setup, include=FALSE}
library(BarrettsProgressionRisk)

#library(biomaRt)
#library(GenomicRanges)
#library(stringr)
#library(ggplot2)
library(ggrepel)
library(GGally)
library(pander)
library(reshape2)
library(gridExtra)
library(CoxHD)
#library(plyr)
#library(dplyr)

source('lib/load_patient_metadata.R')
source('lib/cv-pt-glm.R')
source('lib/common_plots.R')
#source('lib/data_func.R')

data = '~/Data/BarrettsProgressionRisk/'
info.dir = '~/Data/BarrettsProgressionRisk/QDNAseq'

@@

patient.info = read.patient.info(patient.file, demo.file, set='All')$info
head(patient.info)

# validation.patient.info =  subset(all.patient.info, Set == 'Test')
# validation.patient.info = arrange(validation.patient.info, Status, Hospital.Research.ID, Endoscopy.Year, Pathology)
# sum.validation.info = summarise.patient.info(validation.patient.info)

@@

#all.info = rbind(sum.patient.data, sum.validation.info)

@@

pander(
  as.data.frame(sum.patient.data %>% dplyr::summarise(
  'Total Patients'=length(unique(Patient)),
  'Total Samples'=sum(total.samples),
  'Range'=paste(range(end.year-start.year), collapse='-'),
  'Median Years'=paste(median(end.year-start.year),'+/-',round(sd(end.year-start.year),1)),
  'Median Endoscopies'=paste(median(total.endos),'+/-',round(sd(total.endos),1)),
  'Median age'=paste(median(age.diagnosis),'+/-',round(sd(age.diagnosis),1)),
  'gender F:M'=paste(table(gender),collapse=':')
))
, caption='full cohort')
```

cohortSummary = sum.patient.data %>% dplyr::group_by(Status) %>% dplyr::summarise(
  'Total Patients'=length(unique(Patient)),
  'Total Samples'=sum(total.samples),
  'Range follow-up (years)'=paste(range(end.year-start.year), collapse='-'),
  'Mean follow-up (years)'=paste(round(mean(end.year-start.year), 1),'+/-',round(sd(end.year-start.year),1)),
  'Range endoscopies'=paste(range(total.endos), collapse='-'),
  'Mean Endoscopies'=paste(round(mean(total.endos), 1),'+/-',round(sd(total.endos),1)),
  'Mean age'=paste(round(mean(age.diagnosis), 1),'+/-',round(sd(age.diagnosis),1)),
  'gender F:M'=paste(table(gender),collapse=':'),
  'Mean Length'=paste(round(mean(C,na.rm=T), 1),'+/-',round(sd(C,na.rm=T),1)),
  'naC'=length(which(is.na(C)))
)

cohortSummary = full_join(cohortSummary, cbind.data.frame('Status' = c('NP','P'), rbind(
      table(subset(patient.info, Status == 'NP')$Pathology),
      table(subset(patient.info, Status == 'P')$Pathology))), by='Status')

@@

```{r labelsPNP, echo=F, message=F, include=T}
cache.dir = paste(data, 'Analysis/5e6_arms_all', sep='/') # QC'd

@@

pg.sampNOHGD = pg.sampNOHGD %>% filter(!is.na(Prediction))

@@

noHGD.pred.confidence = as.data.frame.matrix(table(pg.sampNOHGD$quants, pg.sampNOHGD$Status))
noHGD.pred.confidence = bind_cols(noHGD.pred.confidence, pg.sampNOHGD %>% dplyr::group_by(quants) %>% dplyr::summarise ( 'mn'=mean(Prediction), 'sd'=sd(Prediction) ))
noHGD.pred.confidence$quant = rownames(noHGD.pred.confidence)

@@

BarrettsProgressionRisk::showPredictionCalibration(noHGD.pred.confidence) + labs(subtitle='HGD excluded model')

@@

  hgd.samples = subset(patient.info, Pathology %in% c('HGD','IMC'))
  hgd.df = dysplasia.df[intersect(hgd.samples$Samplename, rownames(dysplasia.df)),]

  pred = predict(fitNOHGD, newx=hgd.df, s=lambda.opt, type='response')
  colnames(pred) = 'Prediction'
  hgd.samples = base::merge(hgd.samples, pred, by.x='Samplename', by.y='row.names')

  hgd.pred = hgd.samples %>% dplyr::group_by(Patient, Hospital.Research.ID, Status, Endoscopy.Year) %>% 
    dplyr::summarise( max=max(Pathology), pred=Prediction[which.max(Pathology)] )

  rm(no.hgd.plots, coefs, performance.at.1se, models)
}
#hist(hgd.samples$Prediction)

@@

back = back %>% filter(!is.na(Prediction))

@@

ft.fun<-function(NP,P) {
 f = chisq.test(rbind(cbind(NP,P),table(sum.patient.data$Status)))
 cbind.data.frame('p.value'=round(f$p.value, 4))
}

@@

pred.confidence =
  qt %>% as.data.frame.matrix %>% dplyr::group_by(quant) %>% dplyr::mutate( 
  'perc'=P/sum(NP,P), 'p.value'=ft.fun(NP,P)$p.value, 'conf'=ifelse(p.value < 0.05, '*', '') )

@@

pred.confidence = pred.confidence %>% dplyr::mutate(Risk = ifelse(perc<.5, 'Low','High'))
pred.confidence = bind_cols(pred.confidence, 
                            data.frame(ci.low=qbeta(0.025, shape1=pred.confidence$P+.5, shape2 = pred.confidence$NP+.5),
                                      ci.high=qbeta(0.975, shape1=pred.confidence$P+.5, shape2 = pred.confidence$NP+.5)))

pcfit = lm(perc~mn, data=pred.confidence)
rsid = rstandard(pcfit)
pred.confidence[which( rsid >= 1 |  rsid <= -1 ), 'Risk'] = 'Moderate'

@@

```{r pcalib, echo=F, warning=F, message=F, fig.width=8, fig.height=6}
pcal = BarrettsProgressionRisk::showPredictionCalibration(pred.confidence)
pcal
ggsave(filename='plots/pred_calib.png', plot=pcal, width=4, height=4, units='in', dpi=300)
```

@@

```{r relrisk, echo=F, fig.width=6, fig.height=6}
myPal = rev(RColorBrewer::brewer.pal(11, 'RdYlBu'))

unadjRR = ggplot(back, aes(OR)) + geom_histogram(aes(fill=..x..), bins=10, show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(y='n Samples', x='Relative Risk', title='Unadjusted relative risk') + plot.theme
unadjRR

@@

predhist = ggplot(back, aes(Prediction)) + geom_histogram(aes(fill=..x..), breaks=cuts, show.legend = F) +
    scale_fill_gradientn(colors = myPal,  name='') + 
    plot.theme + labs(title='Predictions, all samples model', y='n Samples', x='Probability') 
predhist

ggsave(filename='plots/pred_hist.png', plot=predhist, width=5, height=5, units='in', dpi=300)
```

@@

el = subset(df2, early | late)
el2 = el %>% dplyr::group_by(Status, Patient) %>% dplyr::summarise( diff(arms), diff(CNA) )
wtA = round(wilcox.test(subset(el2, Status == 'Progressor')$`diff(arms)`, subset(el2, Status == 'Non-Progressor')$`diff(arms)`)$p.value, 3)
if (wtA < 0.0001) wtA = '< 0.0001'

@@

```{r jacknife, echo=T, warning=F, message=F, fig.height=10, fig.width=12}
coefs.per.pt = data.frame(matrix(ncol=length(pg.samp), nrow=nrow(ch), dimnames=list( ch$label, names(pg.samp))))
for (pt in names(pg.samp)) {
  lcf = loo.coefs[[pt]][,1]
  names(lcf) = rownames( loo.coefs[[pt]] )
  sharedcfs = intersect(names(lcf), ch$label)
  coefs.per.pt[sharedcfs, pt] = lcf[sharedcfs]
}
coefs.per.pt[is.na(coefs.per.pt)] = 0

nzcoefs = lapply(nzcoefs, function(x) {
  as_tibble(x, rownames='coef')
})

cfs = unique(unlist(sapply(nzcoefs, function(x) x[['coef']])))
loo.coefs = data.frame(matrix(ncol=0,nrow=length(cfs)))
loo.coefs$coef = cfs
for (pt in names(nzcoefs)) {
  loo.coefs = dplyr::full_join(loo.coefs, nzcoefs[[pt]], by='coef')
}
colnames(loo.coefs)[-1] = c(1:length(nzcoefs))
loo.coefs[is.na(loo.coefs)] = 0

ch = as_tibble(coefs$`0.9`[,1,drop=F], rownames='coef')

loo.ch = dplyr::left_join(ch, loo.coefs, by='coef')

@@

ch$jack.se = apply(loo.ch, 1, function(x) {
  jk = bootstrap::jackknife(x[-1], var)
  jk$jack.se
})

@@

# in this case jack.se is the SE of the variance 
# per coefficient
ch$jack.se = apply(coefs.per.pt, 1, function(x) {
  jk = jackknife(x, var)
  jk$jack.se
})
colnames(ch) = c('label','coef','jack.se')
ggplot(ch, aes(reorder(label, coef), coef)) + geom_point() + 
  geom_errorbar(aes(ymin=coef-jack.se, ymax=coef+jack.se)) + 
  coord_flip() +
  plot.theme + labs(x='', title='Jackknifed coefficient CI')

ggplot(ch, aes(reorder(label, coef), coef)) + geom_point() + 
  geom_errorbar(aes(ymin=coef-jack.se*1.96, ymax=coef+jack.se*1.96)) + 
  coord_flip() +
  plot.theme + labs(x='', title='Jackknifed coefficient CI')

@@

ggsave(filename='plots/pred_by_path.png', plot=rp, width=9, height=7, units='in', dpi=300)
```

### 2019-02-05 sWGS QC.Rmd
Updating paper related scripts

dir1 = '~/Data/Ellie/Analysis/multipcf_plots_fitted_perPatient'
dir2 = '~/Data/Ellie/Analysis/VAL_Cohort/sWGS/'

@@

```{r, echo=F}
resVal = read.table(paste(dir2,'resid_var.txt',sep='/'),header=T)
pander(head(resVal[,grep('var1',colnames(resVal))]), caption='Residuals across segments per sample')

@@

setdiff(fc$sample,c(as.character(failedQC$sample), as.character(failedQCVal$sample)))

@@

>>>>

dir1 = '~/Data/Ellie/Analysis/multipcf_plots_fitted_perPatient'
dir2 = '~/Data/Ellie/Analysis/val_test/'

@@

```{r, echo=F}
list.files(dir2

resVal = read.table(paste(dir2,'resid_var.txt',sep='/'),header=T)
pander(head(resVal[,grep('var1',colnames(resVal))]), caption='Residuals across segments per sample')

@@

@@

Using the cutoff `r newcutoff` learned from all of them we would fail `r length(which(resVal$varMAD_median > newcutoff))` samples from the training dataset (Ellie's) and `r length(which(resVal$varMAD_median > newcutoff))` from the initial validation dataset. So while it's less stringent it still fails `r length(which(failedQCVal$varMAD_median > newcutoff))/nrow(resVal)*100`% versus `r length(which(failedQC$varMAD_median > newcutoff))/nrow(res.variance)*100`% of the training dataset.

### 2019-02-05 snp/SNP_Prediction_Analysis.Rmd
Updating paper related scripts

chr.info = get.chr.lengths(file='/tmp/hg19_info.txt')

@@

load('~/Data/Ellie/Analysis/5e6_arms_all-NOQC/model_data.Rdata', verbose = T)
load('~/Data/Ellie/Analysis/5e6_arms_all-NOQC/all.pt.alpha.Rdata', verbose = T)
#alpha = '0'
alpha = '0.9'
fitV = models[[alpha]]
l = performance.at.1se[[alpha]]$lambda
features = coefs[[alpha]]

>>>>

library(BarrettsProgressionRisk)

chr.info = BarrettsProgressionRisk:::chrInfo(build='hg19')

@@

load('~/Data/Ellie/Analysis/5e6_arms_all/model_data.Rdata', verbose = T)
load('~/Data/Ellie/Analysis/5e6_arms_all/all.pt.alpha.Rdata', verbose = T)
#alpha = '0'
alpha = '0.9'
fitV = models[[alpha]]
l = performance.at.1se[[alpha]]$lambda
features = coefs[[alpha]]
